---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Zaremba, Dan"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks,"" and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded.  Points assigned to each item appear in this template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets.** An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question:  How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections ##########################

##### Section 1: (6 points) Summarizing the data.

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)
out <- table(mydata$CLASS,mydata$RINGS)
print(out)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (Looking at the summary of our data set our data types are factor variables, numerical and integers.  The distribution count in Sex between variables I, F, M looks close to one another.  I did notice that we have some large max values in Whole, Shuck, Rings, Volume and Ratio so we probably have some outliers which would result in skewing to the right.)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
table_counts <- table(mydata$SEX, mydata$CLASS)
table_counts 
addmargins(table_counts)  

barplot(table_counts,
        main = "Sex and Class",
        xlab = "Class",
        col = c("red","blue","yellow"), beside = TRUE)

legend("topleft",
       c("Female","Infant","Male"),
       fill = c("red","blue","yellow")
)  

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (As expected the Infants make up the marjority in the lowest Classes A1 and A2, Class is an age classification based on RINGS (A1= youngest,., A6=oldest).  Interesting to see some of the Infants in Classes A4 and A5, perhaps this should be investigated further. )***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(1:nrow(mydata),200,replace=FALSE),] 
plot(work[, 2:6])

```

-----

##### Section 2: (5 points) Summarizing the data using graphics.

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
p <- ggplot(data = mydata, 
            aes(x=VOLUME, y=WHOLE, color=CLASS))+
  geom_point()
p

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
mydata$RATIO2 <- mydata$SHUCK / mydata$WHOLE
MaxShuckWholeRatio <- max(mydata$RATIO2)
MaxShuckWholeRatio

p2 <- ggplot(data = mydata, 
            aes(x=WHOLE, y=SHUCK, color=CLASS))+
  geom_point() 
p2 + geom_abline(intercept = 0, slope = MaxShuckWholeRatio)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (There is more variability in the part(a) data points. A5 is below the other points in part(b) as opposed to part (a). This could mean that weight of the shell is more than the shuck for older classes..)***

-----

##### Section 3: (8 points) Getting insights about the data using graphs.

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
mydataI <- mydata[ which(mydata$SEX=='I'),]
mydataF <- mydata[ which(mydata$SEX=='F'),]
mydataM <- mydata[ which(mydata$SEX=='M'),]
     
par(mfrow = c(3,3)) 
hist(mydataI$RATIO, main="Infant Ratio",col = 'red')
hist(mydataF$RATIO, main="Female Ratio",col = 'blue')
hist(mydataM$RATIO,main="Male Ratio",col = 'yellow')

boxplot(mydataI$RATIO,main="Infant Ratio",col = 'red')
boxplot(mydataF$RATIO,main="Female Ratio",col = 'blue')
boxplot(mydataM$RATIO,main="Male Ratio",col = 'yellow')

qqnorm(mydataI$RATIO,main="Infant Ratio",col = 'red')
qqnorm(mydataF$RATIO,main="Female Ratio",col = 'blue')
qqnorm(mydataM$RATIO,main="Male Ratio",col = 'yellow')

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Looking at the graphs they apear to be normal, Infants look slightly Right-Skewed.)***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
boxplot.stats(mydataI$RATIO)
boxplot.stats(mydataI$RATIO,coef = 3)

boxplot.stats(mydataF$RATIO)
boxplot.stats(mydataF$RATIO,coef = 3)

boxplot.stats(mydataM$RATIO)
boxplot.stats(mydataM$RATIO,coef = 3)

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (The Infants Ratio has 8 outliers. 0.2693371, 0.2218308, 0.2403394, 0.2263294, 0.2249577, 0.2300704, 0.2290478, and 0.2232339.
The Female Ratio has 6 outliers. 0.31176204, 0.21216140, 0.21465603, 0.21306058, 0.23497668, and 0.06733877.
The Male Ratio has 5 outliers. 0.2609861, 0.2378764, 0.2345924, 0.2356349, and 0.2286735.
The Infant and Female Ratios are the only ones with extreme outliers.
The Infant extreme outlier being 0.2693371 and the female extreme outlier being 0.311762
)***

-----

##### Section 4: (8 points) Getting insights about possible predictors.

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
t<-ggplot(data = mydata, aes(x=CLASS, y=VOLUME)) + geom_boxplot(aes(fill=CLASS))
r<-ggplot(data = mydata, aes(x=CLASS, y=WHOLE)) + geom_boxplot(aes(fill=CLASS))
s1 <- ggplot(data = mydata,aes(x=RINGS, y=VOLUME, color=RINGS))+ geom_point()
s2 <- ggplot(data = mydata,aes(x=RINGS, y=WHOLE, color=RINGS))+ geom_point()
grid.arrange(t, r, s1, s2, ncol=2)

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (I believe the variables can perform as predictors of age as reflected in the upward trend that we see in the boxplots. But refinement is needed as we noticed some Infants bucketed in the Class 4 and 5)***

-----

##### Section 5: (12 points) Getting insights regarding different groups in the data.

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
VolumeM <- aggregate(mydata$VOLUME, by = list(mydata$SEX, mydata$CLASS), mean)
matrix(VolumeM$x,nrow=3, ncol = 5, byrow = FALSE, dimnames = list(c("Female","Infant","Male"), c("A1","A2","A3","A4","A5")))

ShuckM <- aggregate(mydata$SHUCK, by = list(mydata$SEX, mydata$CLASS), mean)
matrix(ShuckM$x,nrow=3, ncol = 5, byrow = FALSE, dimnames = list(c("Female","Infant","Male"), c("A1","A2","A3","A4","A5")))

RatioM <- aggregate(mydata$RATIO, by = list(mydata$SEX, mydata$CLASS), mean)
matrix(RatioM$x,nrow=3, ncol = 5, byrow = FALSE, dimnames = list(c("Female","Infant","Male"), c("A1","A2","A3","A4","A5")))

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
interaction.plot(x.factor = mydata$CLASS,    # variable to plot on x-axis
                 trace.factor = mydata$SEX, # variable to specify "traces"; here, lines
                 response = mydata$RATIO,    # variable to plot on y-axis
                 fun = mean,  # summary statistic to be plotted for response variable
                 type = "l",     # type of plot, here "l" for lines
                 ylab = "Ratio",
                 xlab = "Class",
                 col = c("red4", "blue4","yellow4"),
                 lty = 1,  # line type
                 lwd = 2,  # line width
                 trace.label = "Sex",  # label for legend
                 xpd = FALSE) #,  # 'clip' legend at border

interaction.plot(x.factor = mydata$CLASS,    # variable to plot on x-axis
                 trace.factor = mydata$SEX, # variable to specify "traces"; here, lines
                 response = mydata$VOLUME,    # variable to plot on y-axis
                 fun = mean,  # summary statistic to be plotted for response variable
                 type = "l",     # type of plot, here "l" for lines
                 ylab = "Volume",
                 xlab = "Class",
                 col = c("red4", "blue4","yellow4"),
                 lty = 1,  # line type
                 lwd = 2,  # line width
                 trace.label = "Sex",  # label for legend
                 xpd = FALSE) #,  # 'clip' legend at border

interaction.plot(x.factor = mydata$CLASS,    # variable to plot on x-axis
                 trace.factor = mydata$SEX, # variable to specify "traces"; here, lines
                 response = mydata$SHUCK,    # variable to plot on y-axis
                 fun = mean,  # summary statistic to be plotted for response variable
                 type = "l",     # type of plot, here "l" for lines
                 ylab = "Shuck",
                 xlab = "Class",
                 col = c("red4", "blue4","yellow4"),
                 lty = 1,  # line type
                 lwd = 2,  # line width
                 trace.label = "Sex",  # label for legend
                 xpd = FALSE) #,  # 'clip' legend at border

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (One question I have is that I notice that the highest mean Shuck is in Class A4, why does it decrease in A5?)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
mydataI2 <- mydata[mydata$SEX =='I' & mydata$RINGS < 16,]
mydataA <- mydata[mydata$SEX !='I' & mydata$RINGS < 16,]

v<-ggplot(data = mydataI2, aes(x=RINGS, y=VOLUME)) + geom_boxplot(aes(fill=RINGS))
w<-ggplot(data = mydataA, aes(x=RINGS, y=VOLUME)) + geom_boxplot(aes(fill=RINGS))
x<-ggplot(data = mydataI2, aes(x=RINGS, y=WHOLE)) + geom_boxplot(aes(fill=RINGS))
y<-ggplot(data = mydataA, aes(x=RINGS, y=WHOLE)) + geom_boxplot(aes(fill=RINGS))

grid.arrange(v, w, x, y, ncol=2)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (Enter your answer here.)***

-----

##### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).

**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (Skewness was witnessed in the data, best to have nonrmalized data, also the factors seem highly correlated, which can cause issues when modeling***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (I would want to know all I could about how the sample was performed and about the numbers reflected in it.  I would want to know what was our sample population, how did we get to this numbered sample? what was the process?, where did the sample take place? Was the sample random? Was there any bias? )***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Some concerns that I have is that their may be error or bias, human error in reporting/measuring or bias in selecting the location for the sample. Causality can’t be determined. There is no control to compare the data. Although these studies can helpful in identifying correlations that may exist between sex, class, weight etc. )***
